Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-11 paper/manuscript.tex
Edits to beginning of the document

The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that do not exist in as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly. 

@@

\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}

@@

The calculations use non-physical intermediate states where atoms are changed from one type to another or can be removed or inserted by turning on or off the interactions with their environment. 

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with AFE methods. In these types of simulations so-called \textit{alchemical states} permit converged calculations for large changes, such as the entire removal of a ligand from a binding site, as seen in fig.~\ref{fig:fig_what_is_alchemy} (E). This type of simulation is known as an absolute AFE calculation. Other typical problems tackled with AFE calculations can be as simple as conformational changes of a molecule that exhibit a high barrier between these (fig.~\ref{fig:fig_what_is_alchemy} (A)); computation of partition coefficients, such as $\log P$ or $\log D$ (fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded}; and membrane insertions (fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}. Then there are common types of AFE that estimate free energies of binding: the ligand stays the same, but a protein residue is alchemically mutated and the effect on the binding affinity is probed (fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate}, changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}, absolute binding free energies (fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{}, and relative free energies of binding where the difference of binding free energy between to similar ligands is evaluated (fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{}.
In order to obtain a free energy from these AFE simulations, historically two common approaches were used for the analysis: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or Bennett's acceptance ratio~\cite{bennett1976efficient} and thermodynamic integration (TI) for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. This foundational work, combined with the methodological, technological, and hardware improvements of the last 5-10 years have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.

@@

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in fig.~\ref{fig:fig_topology}). In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations. In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.

@@

The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g. consider a transformation of benzene (A) to benzyl alcohol (B) fig.~\ref{fig:topology}.

@@

Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of fig.~\ref{fig:fig_topology}. 

@@

In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in fig.~\ref{fig:fig_types_of_networks} (A) is used.. However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in fig.~\ref{fig:fig_types_of_networks}.

@@

\subsubsection{What different sampling schemes exist and which one should I use?}
With all AFE simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used. The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see fig.~\ref{fig:fig_sampling_scheme}(A). This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}. A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica. This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}. Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in fig.~\ref{fig:fig_sampling_scheme}(B). It is supported in various software packages that provide AFE implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}. A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambdas$, as seen in fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}. The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations. A schematic of this approach is shown in fig.~\ref{fig:fig_sampling_scheme}. 
\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states. \textbf{B}: Hamiltonian replica exchange scheme. \textbf{C}: Single replica scheme sampling from all $\lambda$ states. \textbf{D}: Non-equilibrium sampling scheme.}
    \label{fig:fig_sampling_scheme}
\end{figure} 
The best practice recommendation, is to use replica exchange type sampling schemes (fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (fig.~\ref{fig:fig_sampling_scheme}(A)). Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.

>>>>

The hallmark of these methods is the use of "bridging" potential energy functions representing \emph{alchemical} intermediate states that cannot exist in as real chemical species. The data collected from these bridging alchemical thermodynamic states allows the efficient computation of transfer free energies (or differences in transfer free energies) with orders of magnitude less simulation time than simulating the transfer process directly. 

@@

%\todo[inline, color=green!20]{JDC: Should we migrate this document to \url{https://github.com/alchemistry}? ASJSM: Let's migrate with the submitted version, but leave it here for now?}

@@

\todo[inline, color={yellow!20}]{JDC: Do we really want to introduce the abbreviation "AFE" when we can just use the word "alchemical" wherever it might be needed?}
These calculations use non-physical\footnote{Here, the non-physical nature of the transformation is referred to as "alchemical", a term coined by Tembre and McCammon in Ref.~\cite{tembre1984ligand}.} intermediate states in which the chemical identity of some portion of the system (such as a small molecule ligand or receptor sidechain) is changed by modifying the potential governing the interaction of the atoms being modified, inserted, or deleted with their environment. 

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with AFE methods. 
\todo[inline, color={yellow!20}]{JDC: Not sure what is meant by "conventional computational methods". Perhaps we can be precise and say "unbiased molecular dynamics simulations"?}
In AFE simulations, the introduction of intermediate \textit{alchemical states} that bridge the high-probability regions of configuration space between the two physical endstates of interest permits the free energy of large transformations to be robustly computed.
%such as the entire removal of a ligand from a binding site (and \emph{absolute AFE calculation}), as illustrated in Fig.~\ref{fig:fig_what_is_alchemy} (E). 
AFE calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A)); 
\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations?}
computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}. 
AFE calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2016accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

Once an AFE simulation as been performed, the data must be analyzed to estimate the free energy for the transformation of interest.
Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. 
More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{mbar} or removing the need for global equilibrium~\cite{tram}
\todo[inline, color={yellow!20}]{JDC: The key pivotal moment was the Shirts paper on extremely precise free energy calculations, which demonstrated that a combination of technologies can produce robust estimates with useful precision.}
Important studies continued during the 2000's often leading to improved implementation of the methods in common simulation software~\cite{vanderspoel2005gromacs, mermelstein2018fast, wang2015accurate, hedges2019biosimspace}. 
This foundational work, combined with the methodological, technological, and hardware improvements of the last 5--10 years, have led to an explosion of interest and direct commercial application of the technologies~\cite{wang2015accurate, fratev2019improved, schindler2020largescale, cournia2017relative, sherborne2016collaborating}.

@@

An additional consideration arises for single topology relative free energy calculations. In this class of alchemical free energy calculations it is necessary to generate a molecular topology that may describe the initial and final states of the perturbation (see different topologies as seen in Fig.~\ref{fig:fig_topology}). In cases where the end-states have high topological similarity and high structural overlap this is relatively straightforward and typically handled by use of maximum common substructure calculations. In situations where the end-state topologies differ significantly, or there is relatively little spatial overlap between the two end-states some user intervention may be necessary to produce a satisfactory input topology.

@@

The distinction between these can be illustrated by considering a hypothetical transformation from molecule A to molecule B, where both atoms share a common substructure but differ in their substituents; e.g. consider a transformation of benzene (A) to benzyl alcohol (B) Fig.~\ref{fig:topology}.

@@

Thus, in a single topology calculation, atoms may change their type so relatively few dummy atoms are created. This is illustrated in left arm of Fig.~\ref{fig:fig_topology}. 

@@

In some cases, during the lead optimization stage, or if perturbations would be rather large a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} (A) is used.. However, adding redundancy into the network means that a better error analysis can be carried out, by looking at cycle closure errors as discussed in sec.~\ref{subsec:predictions}, with an example given in Fig.~\ref{fig:fig_types_of_networks}.

@@

\subsubsection{What different sampling schemes exist and which one should I use?}
With all AFE simulations having the requirement of sampling from different sets of $\lambda$ states in common different approaches have been designed. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes used. The simplest approach is defining a set of fixed $\lambda$ values and then run independent simulations at each of the intermediate $\lambda$ values, see Fig.~\ref{fig:fig_sampling_scheme}(A). This type of scheme is currently used for AMBER TI calculations~\cite{song2019} and for BioSimSpace~\cite{hedges2019biosimspace}. A slightly more complex approach is allowing replica mixing, meaning that each independent $\lambda$ simulation is attempted to be exchanged with a neighbouring $\lambda$ replica. This is based on ideas of Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{wendsen1986replica}. Using a Metropolis Hastings based acceptance criterion for exchanges means that the generated ensemble of all replicas is still sampling from the Boltzmann distribution and has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}(B). It is supported in various software packages that provide AFE implementations, such as GROMACS~\cite{aldeghi2015accurate}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}. A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\lambda$ space by optimally working out weights that allow switching between different intermediate $\lambda$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}(C) and also referred to as self adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in~\cite{andrearizzi2019choderalab}. The last approach available in conjunction with GROMACS makes use of non-equilibrium simulations~\cite{aldeghi2018accurate}. In this approach, only end state $\lambda$ replicas ($\lambda$=0, $\lambda=1$) are simulated at equilibrium intermediate information is generated from non-equilibrium simulations. A schematic of this approach is shown in Fig.~\ref{fig:fig_sampling_scheme}. 
\ref{fig:fig_sampling_scheme} A:\cite{hedges2019biosimspace, song2019using } B\cite{aldeghi2015accurate, wang2015accurate,jiang2019computing} C:\cite{li2007simulated, tan2017optimally} D:\cite{aldeghi2018accurate}
\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig8_sampl_scheme/Figure.pdf}
    \caption{\textbf{Four sampling strategies available in different software packages.} \textbf{A}: Multiple replicas in parallel at different lambda states. \textbf{B}: Hamiltonian replica exchange scheme. \textbf{C}: Single replica scheme sampling from all $\lambda$ states. \textbf{D}: Non-equilibrium sampling scheme.}
    \label{fig:fig_sampling_scheme}
\end{figure} 
The best practice recommendation, is to use replica exchange type sampling schemes (Fig.~\ref{fig:fig_sampling_scheme}(B)), or in absence of their availability and when looking at systems with fast mixing times parallel replica schemes are acceptable (Fig.~\ref{fig:fig_sampling_scheme}(A)). Single replica schemes and non-equilibrium schemes are not as established yet, but are very promising.
